An assessment of statistical guidance for authors at influential academic journals across scientific disciplines: Preliminary analysis

### 2021-08-16 analysis/preliminaryAnalysis.Rmd
Update preliminaryAnalysis.Rmd
added new graphs and other assorted changes

---
title: "An assessment of statistical guidance for authors at influential academic journals across scientific disciplines: Preliminary analysis"
author: "Analysis performed by: Maia Salholz-Hillel & Tom Hardwicke"
date: "Last updated: `r format(Sys.time(), '%d %B, %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

@@

```{r load_packages}
library(knitr) # for literate programming
library(papaja) # for article template
library(tidyverse) # for data munging
library(tidylog) # inline analysis feedback
library(here) # for finding files
library(gtsummary) # for summary tables
```

@@

```{r load_processed_data}
# loads the processed data files output by the preprocessing performed above
load(here('data','processed','d_coding.rds'))
load(here('data','processed','d_journals.rds'))
load(here('data','processed','lookup_external_guidance.rds'))
```

```{r set_style_defaults}
theme_gtsummary_compact() # set table theme
```

## Introduction
<br>
We have now completed data collection for the first phase of this project. We extracted author-facing statistical guidance provided by the top-15 journals (by Impact Factor) in each of 22 scientific disciplines (i.e., total N = 330 journals). The preregistered study protocol is available [here](https://osf.io/cz9g3/). Below we present a preliminary analysis of the phase one data and highlight some issues to discuss as a team.

## Method summary

@@

**Note** Some publishers provided statistical guidance that was shared across their portfolio of journals (specifically, 32 Nature journals, 12 Cell journals, and 2 Frontiers journals) . Publisher guidance was coded by two team members (as above), but is represented in our data and analysis for each individual journal the guidance applies to. For example, the Nature guidance is inherited by 32 journals so appears 32 times in the dataset.

## Results

Look at counts and proportions mentioning each topic. Use all journals for denominator. Overall and by field.

``````{r tbl_counts_props}
d_journals %>% 
  select(journal, esi_field) %>% 
  left_join(d_coding, ., by = "journal") %>%

@@

  select(esi_field, starts_with("has_")) %>% 
  gtsummary::tbl_summary(
    by = esi_field,
      # statistic = list(everything() ~ "{p}% ({n} / {N})")
  ) %>%
  add_overall %>% 
  modify_footnote(everything() ~ NA) %>% 
  modify_caption("**Number and proportion of journals overall and by field providing internal statistical guidance by topic.** Denominator for all proportions is the number of journals in the field, regardless of whether the journal has internal guidance.")
```

@@

Try a mosaic plot for internal guidance...nope!

@@

d_internal_guidance <-
  d_coding %>% 
  select(journal, starts_with("has_") & (!contains("guidance"))) %>% 
  pivot_longer(
    -journal,
    names_to = "topic",
    values_to = "outcome",
    values_drop_na = FALSE
  ) %>% 
  mutate(outcome = case_when(
    outcome ~ "Mentioned",
    !outcome ~ "Not mentioned",
    TRUE ~ "Not applicable (No internal guidance)"
  ))

# plt <- 
  d_internal_guidance %>%
  ggplot(aes(x = topic, y = factor(journal))) +
  geom_tile(aes(fill = outcome), colour = 'black', size = .25) +
  scale_fill_manual(
    values = my_palette,
    # labels = c('Not mentioned', 'Unclear', 'Mentioned')
  ) +
  scale_x_discrete(position = 'top') +
  papaja::theme_apa() +
  xlab("Topics mentioned") +
  theme(plot.margin = margin(0, 50, 0, 0),
        axis.ticks = element_blank(),
        axis.title.y =element_blank(),
        panel.grid.major=element_line(colour=NA), 
        panel.grid.minor=element_line(colour=NA),
        legend.position = 'bottom',
        axis.text.x  = element_text(angle = 45, hjust = 0),
        strip.background = element_rect(
          color="black", fill="white", size=0, linetype="solid"
        ),
        strip.text.y = element_text(
          size = 8, color = "black", face = 'bold'
        ))

@@

Try bar chart

```{r plot_internal_guidance_bar}
d_internal_guidance %>% 
  ggplot(aes(y = topic, fill = outcome)) +
  geom_bar() +
  scale_fill_manual(values = my_palette)
```

Look at external guidance

@@

## Issues to address

@@

>>>>

---
title: "An assessment of statistical guidance for authors at influential academic journals across scientific disciplines: Preliminary analysis"
author: "Analysis performed by: Maia Salholz-Hillel & Tom Hardwicke"
date: "Last updated: `r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::html_document2
editor_options: 
  chunk_output_type: console
---

@@

```{r load_packages}
library(knitr) # for literate programming
library(papaja) # for article template
library(tidyverse) # for data munging
library(tidylog) # inline analysis feedback
library(here) # for finding files
library(gtsummary) # for summary tables
library(lemon) # to have axes across all graph facets
library(ggthemr) # for ggplot theme
library(ggrepel) # for ggplot text labels
```

@@

```{r load_processed_data}
# loads the processed data files output by the preprocessing performed above
load(here('data','processed','d_all.rds'))
load(here('data','processed','lookup_external_guidance.rds'))
```

```{r set_style_defaults}
theme_gtsummary_compact() # set table theme
ggthemr('fresh') # set ggplot theme
```

# Introduction
<br>
We have now completed data collection for the first phase of this project. We extracted author-facing statistical guidance provided by the top-15 journals (by Impact Factor) in each of 22 scientific disciplines (i.e., total N = 330 journals). The preregistered study protocol is available [here](https://osf.io/cz9g3/). Below we present a preliminary analysis of the phase one data and highlight some issues to discuss as a team.

# Methods summary

@@

```{r shared_publisher_journals}
# count number of journals that have shared publisher guidance
publisher_nature_n <- d_all %>% filter(publisher_guidance == "Nature") %>% nrow()
publisher_cell_n <- d_all %>% filter(publisher_guidance == "Cell") %>% nrow()
publisher_frontiers_n <- d_all %>% filter(publisher_guidance == "Frontiers") %>% nrow()
```

**Note:** Some publishers provided statistical guidance that was shared across their portfolio of journals (specifically, `r publisher_nature_n` Nature journals, `r publisher_cell_n` Cell journals, and `r publisher_frontiers_n` Frontiers journals) . Publisher guidance was coded by two team members (as above), but is represented in our data and analysis for each individual journal the guidance applies to. For example, the Nature guidance is inherited by 32 journals so appears 32 times in the dataset.

# Results

```{r summarise_data}

# couldn't figure out to extract summary data from gtsummary::tblsummary so making my own summary dataframe
sum_tab <- d_all %>%

@@

  select(esi_field, starts_with("has_")) %>% 
  group_by(esi_field) %>%
  summarise(across(starts_with("has_"), list(n = sum, N = ~n()), .names = "{col}__{fn}")) %>%
  adorn_totals(name = 'OVERALL') %>%
  pivot_longer(cols = !esi_field, names_to = c('variable', '.value'), names_sep = '__', names_prefix = 'has_') %>%
  mutate(prop = n/N, percent = round(prop*100, 0)) %>%
  mutate(esi_abbr = case_when( # creating abbreivated field labels
    esi_field == "OVERALL" ~ "OVERALL",
    esi_field == "AGRICULTURAL SCIENCES" ~ "AGRI",
    esi_field == "BIOLOGY & BIOCHEMISTRY" ~ "BIO",       
    esi_field == "CHEMISTRY" ~ "CHEM" ,                   
    esi_field == "CLINICAL MEDICINE" ~ "MED",           
    esi_field == "COMPUTER SCIENCE" ~ "COMSCI",             
    esi_field == "ECONOMICS & BUSINESS" ~ "ECON",         
    esi_field == "ENGINEERING" ~ "ENGIN",                  
    esi_field == "ENVIRONMENT_ECOLOGY" ~ "ECO",         
    esi_field == "GEOSCIENCES" ~ "GEO",                  
    esi_field == "IMMUNOLOGY" ~ "IMMUN",                   
    esi_field == "MATERIALS SCIENCE" ~ "MATSCI",            
    esi_field == "MATHEMATICS" ~ "MATH",                
    esi_field == "MICROBIOLOGY" ~ "MICBIO",                 
    esi_field == "MOLECULAR BIOLOGY & GENETICS" ~ "MOLBIO", 
    esi_field == "Multidisciplinary" ~ "MULTI",            
    esi_field == "NEUROSCIENCE & BEHAVIOR" ~ "NEURO",     
    esi_field == "PHARMACOLOGY & TOXICOLOGY" ~ "PHARM",    
    esi_field == "PHYSICS" ~ "PHYS",                     
    esi_field == "PLANT & ANIMAL SCIENCE" ~ "PLANT",       
    esi_field == "PSYCHIATRY_PSYCHOLOGY" ~ "PSY",       
    esi_field == "SOCIAL SCIENCES, GENERAL" ~ "SOCSCI",     
    esi_field == "SPACE SCIENCE" ~ "SPACE"          
  )) %>%
  mutate(esi_field_with_abbr = paste0(esi_field,' (', esi_abbr,')'))

# get fields with no statistical guidance at all
null_fields <- sum_tab %>% 
  filter(
    variable == "guidance",
    prop == 0
  ) %>%
  pull(esi_field)

# select only data on general any/internal/external statistical guidance classifications 
sum_tab_general <- sum_tab %>%
  filter(variable %in% c('guidance', 'internal_guidance', 'internal_guidance_section', 'external_guidance')) %>% # select only data on general classifications
  mutate(variable = factor(variable, levels = c('guidance', 'internal_guidance', 'internal_guidance_section', 'external_guidance'))) # set variable order

# select only data on topics
sum_tab_topics <- sum_tab %>%
  filter(esi_field %notin% null_fields) %>% # remove fields that have no statistical guidance at all
  filter(variable %notin% c('guidance', 'internal_guidance', 'internal_guidance_section', 'external_guidance')) %>% # select only data on topics
  rename(topic = variable)

# get list of statistical topics in order of overall proportions
ordered_topics <- sum_tab_topics %>%
  filter(esi_field == "OVERALL") %>%
  arrange(desc(prop)) %>% # order from highest to lowest
  pull(topic)

# set order of topics
sum_tab_topics <- sum_tab_topics %>%
  mutate(topic = factor(topic, levels = ordered_topics))
```

Figure 1 shows the percentage of journals in each scientific discipline (N = 15) and overall (N = 330) that offered (a) any statistical guidance (`r sum_tab %>% filter(esi_field == "OVERALL", variable == "guidance") %>% pull(percent)`% overall); (b) journal-specific statistical guidance (`r sum_tab %>% filter(esi_field == "OVERALL", variable == "internal_guidance") %>% pull(percent)`% overall); (c) a dedicated statistical guidance section (`r sum_tab %>% filter(esi_field == "OVERALL", variable == "internal_guidance_section") %>% pull(percent)`% overall); or referred to (d) external guidelines, such as reporting guidelines or academic articles (`r sum_tab %>% filter(esi_field == "OVERALL", variable == "external_guidance") %>% pull(percent)`% overall).

How much guidance was provided? Some journals mention just one thing, some mention several.

@@

<br>

```{r fig.cap = "Percentage of journals offering statistical guidance by scientific discipline. Overall proportion shown by black diamond. Graphs ordered from left to right and top to bottom in order of highest proportion overall across scientific disciplines.", fig.topcaption = TRUE, fig.width = 10, fig.height = 8}
# extract overall data
sum_tab_general_overall <- sum_tab_general %>%
  filter(esi_field == "OVERALL")

sum_tab_general %>%
  filter(esi_field != "OVERALL") %>%
  ggplot(aes(x=percent, y = 1, colour = str_wrap(esi_field_with_abbr,25))) +
  facet_rep_wrap(facets = vars(variable), ncol = 2, nrow = 2) +
  coord_capped_cart(bottom='both', left='both') + # to break axes in bottom left corner
  geom_point(size = 5, position = position_stack(), alpha = .5) +
  geom_point(size = 6, position = position_stack(), data = sum_tab_general_overall, colour = 'black', shape = 18) +
  geom_text_repel(
    aes(label = ifelse(percent > 0,esi_abbr,''),
        segment.square = T,
        segment.inflect = T), #
    #point.padding = 0.2,
    force_pull   = 0, # do not pull toward data points
    nudge_y      = 15,
    #nudge_x = 5,
    direction    = "x",
    angle        = 90,
    hjust        = 0,
    segment.size = 0.2,
    segment.curvature = -1e-20,
    max.iter = 1e4, max.time = 1,
    max.overlaps = Inf, # ensure all labels are shown even if overlapping
    show.legend = FALSE # do not add legend entry
  ) +
  scale_y_continuous(limits = c(0,15), name = 'number of scientific disciplines') +
  xlim(0,100) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(),
        legend.position = 'bottom') +
  scale_colour_discrete(name = 'Scientific disciplines') +
  guides(colour = guide_legend(nrow = 6, title.position = 'top'))
```

```{r fig.cap = "Percentage of journals offering statistical guidance on twenty key topics by scientific discipline. Overall proportion shown by black diamond. For presentational purposes, disciplines offering no guidance on individual topics are shown, but not labelled. Two scientific disciplines in which no journals offered any statistical guidance at all are not shown. Graphs ordered from left to right and top to bottom in order of highest proportion overall across scientific disciplines.", fig.topcaption = TRUE, fig.width = 10, fig.height = 28}
# extract overall data
sum_tab_topics_overall <- sum_tab_topics %>%
  filter(esi_field == "OVERALL")

sum_tab_topics %>%
  filter(esi_field != "OVERALL") %>%
  ggplot(aes(x=percent, y = 1, colour = str_wrap(esi_field_with_abbr,25))) +
  facet_rep_wrap(facets = vars(topic), ncol = 2) +
  coord_capped_cart(bottom='both', left='both') + # to break axes in bottom left corner
  geom_point(size = 5, position = position_stack(), alpha = .5) +
  geom_point(size = 6, position = position_stack(), data = sum_tab_topics_overall, colour = 'black', shape = 18) +
  geom_text_repel(
    aes(label = ifelse(percent > 0,esi_abbr,''),
        segment.square = T,
        segment.inflect = T), #
    #point.padding = 0.2,
    force_pull   = 0, # do not pull toward data points
    nudge_y      = 25,
    nudge_x = 10,
    direction    = "x",
    angle        = 90,
    hjust        = 0,
    segment.size = 0.2,
    segment.curvature = -1e-20,
    max.iter = 1e4, max.time = 1,
    max.overlaps = Inf, # ensure all labels are shown even if overlapping
    show.legend = FALSE # do not add legend entry
  ) +
  scale_y_continuous(limits = c(0,25), name = 'number of scientific disciplines') +
  xlim(0,100) +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(),
        legend.position = 'bottom') +
  scale_colour_discrete(name = 'Scientific disciplines') +
  guides(colour = guide_legend(nrow = 5, title.position = 'top'))
```

```{r}
# try binning method
# forPlot %>%
#   #mutate(bin = cut_width(prop, width = 0.2, center = 0.1)) %>%
#   mutate(bin = cut_width(percent, width = 20, center = 10)) %>%
#   group_by(esi_field,bin) %>%
#   summarise(n = n()) %>%
#   group_by(bin) %>%
#   summarise(esi_field = esi_field, pos = cumsum(n)) %>%
#   merge(forPlot, by = "esi_field") %>% # rejoin columns from original dataframe
#   ggplot(aes(x=percent, y = pos, colour = esi_field)) +
#   geom_vline(xintercept = sum_tab %>% filter(variable == "guidance", esi_field == "OVERALL") %>% pull(percent)) +
#   geom_point(size = 5) +
#   geom_label_repel(aes(label = esi_abbr)) +
#   theme_minimal() +
#   theme(axis.title.y = element_blank(),
#         axis.text.y = element_blank(),
#         axis.ticks.y = element_blank()) +
#   scale_colour_discrete()
```

```{r}
# # trying out a dot histogram
# d <- sum_tab %>%
#   #filter(variable %in% c('guidance', 'internal_guidance', 'internal_guidance_section', 'external_guidance')) %>%
#   filter(variable %in% c('guidance'))
# 
# sum_tab %>%
#   filter(variable %in% c('guidance', 'internal_guidance', 'internal_guidance_section', 'external_guidance')) %>%
#   ggplot(aes(x = prop, fill=esi_field, label = esi_abbr, group = esi_field)) + 
#     #geom_histogram(aes(fill=esi_field), binwidth=0.1, colour="grey20", lwd=0.2) +
#     geom_dotplot(binwidth=0.1, colour="grey20", lwd=0.2, stackgroups = T, binpositions = 'all', method = 'histodot', dotsize = 0.25, stackdir = 'center') +
#     facet_grid(rows = vars(variable)) +
#     theme(axis.title.y = element_blank(), # y axis labels are meaningless on this plot, remove them
#       axis.text.y = element_blank(),
#       legend.position = 'bottom')
# 
#   # looks ok... but difficult to tell which field is which (too many similar colours, and have to keep looking at legend)
#   # could we put the field abbreviation text inside the dots? (maybe shorten to two letters)
#   # also for OVERALL we need something different, e.g., a black vertical line
```

@@

d_internal_guidance_noPublisher <-
  d_all %>%
  filter(has_internal_guidance == T, # select journals with internal guidance
         has_publisher_guidance == F | journal == "Scientific Data") # remove journals with shared publisher guidance except for Scientific Data (which shared guidance and its own guidance)

# were going to represent journals with publisher guidance in a single row per publisher. 
# select exemplar journals for each publisher
publisher_nature <- d_all %>% filter(journal == "Nature") %>% mutate(journal = paste0("Nature journals"," (n=", publisher_nature_n,')'))
publisher_cell <- d_all %>% filter(journal == "CELL") %>% mutate(journal = paste0("Cell journals"," (n=", publisher_cell_n,')'))
publiher_frontiers <- d_all %>% filter(journal == "Frontiers in Microbiology") %>% mutate(journal = paste0("Frontiers journals"," (n=", publisher_frontiers_n,')'))

# combine the journals with no publisher guidance with the publisher guidance rows
d_internal_guidance <- bind_rows(d_internal_guidance_noPublisher, publisher_nature, publisher_cell, publiher_frontiers)

# for journals with shared publisher guidance we only need only row

plt <- d_internal_guidance %>%
  select(journal, starts_with("has_") & (!contains("guidance"))) %>%
  pivot_longer(
    -journal,
    names_to = "topic",
    values_to = "outcome",
    values_drop_na = FALSE
  ) %>%
  mutate(outcome = case_when(
    outcome ~ "Mentioned",
    !outcome ~ "Not mentioned",
    TRUE ~ "Not applicable (No internal guidance)"
  )) %>%
  ggplot(aes(x = topic, y = str_wrap(journal,25))) +
    geom_tile(aes(fill = outcome), colour = 'black', size = .25) +
    scale_fill_manual(
      values = my_palette,
      # labels = c('Not mentioned', 'Unclear', 'Mentioned')
    ) +
    scale_x_discrete(position = 'top') +
    #papaja::theme_apa() +
    xlab("Topics mentioned") +
    theme(plot.margin = margin(0, 50, 0, 0),
          axis.ticks = element_blank(),
          axis.title.y =element_blank(),
          axis.line = element_blank(),
          panel.grid.major=element_line(colour=NA),
          panel.grid.minor=element_line(colour=NA),
          legend.position = 'bottom',
          axis.text.x  = element_text(angle = 45, hjust = 0),
          strip.background = element_rect(
            color="black", fill="white", size=0, linetype="solid"
          ),
          strip.text.y = element_text(
            size = 8, color = "black", face = 'bold'
          ))

@@

```{r fig.height = 40}
plt
```
Look at external guidance

@@

# Issues to address

@@

# Appendix 1

Tabular data (represented in Figures 1 and 2)
Look at counts and proportions mentioning each topic. Use all journals for denominator. Overall and by field.

```{r tbl_counts_props}
d_all %>%

  select(esi_field, starts_with("has_")) %>% 
  gtsummary::tbl_summary(
    by = esi_field,
      # statistic = list(everything() ~ "{p}% ({n} / {N})")
  ) %>%
  add_overall %>% 
  modify_footnote(everything() ~ NA) %>% 
  modify_caption("**Number and percentage of journals overall and by scientific discipline providing internal statistical guidance by topic.** Denominator for all proportions is the number of journals in the field, regardless of whether the journal has internal guidance.")
```
